Validation of Visual Statistical Inference, Applied to Linear Models

### 2012-05-25 paper_jasa.tex
power and lineup assessment

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\\hline
1 &  1 & 0.0500 & 2 &  1 & 0.0975 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0001 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &        &   &   &        & 4 & 4 & 0.0000 & 5 & 4 & 0.0000 \\\hline
\end{tabular}
\end{center}
\label{pvalue}
\end{table}

@@

\red{Note that $F_{X, \theta}$ is the distribution of $X$ under the alternative with given parameter value $\theta$.}

@@

\red{The following considerations deal with assessing the `difficulty' of  a lineup of size $m$. I think it's very relevant, but I don't see a natural equivalent in the classical testing situation.}

Suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$, the classical test statistic. Suppose the associated test statistic in classical test is observed as  $t_{obs}$ with $p$-value $p_D$. By definition we have

@@

%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$, the classical test statistic.
%, with non-centrality parameter $\delta$. 
Then the distribution function of $p$-value $p_D$ under $H_0$ is uniform, since:  
\begin{eqnarray}\label{dist_p}
F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
  &=& 1-Pr\left(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) \right) \nonumber \\
  &=& 1-Pr\left(|t_{obs}| \le F_{|t|}^{-1}(1-p)\right) \nonumber \\
  &=&%\left\{ \begin{array}{ll}
          1-F_{|t|}\left( F_{|t|}^{-1}(1-p)\right)=p \mbox{ ; under $H_0$} 
 %         1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%       \end{array} \right.     
\end{eqnarray}

@@

\blue{Similarly, we will assume that $p_{0,i}$, $i=1, ..., m-1$ are the $p$-values associated with data corresponding to the $m-1$ null plots. Since this data is generated consistently with the null hypothesis,  the $p$-values follow a standard Uniform distribution, $p_{i,0} \sim U[0,1], i= 1, ..., m-1$ (by equation \ref{dist_p}). }

@@

With that, we define $X$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.

Then $X \sim B_{m-1, E[p_D]}$, and the probability that an onlooker will pick the data plot in a given lineup is 
\[
P(X=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
\]

@@

Figure \ref{fig:pval_power} gives an overview of the probability of picking the data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. We see that for a $p$-value $p_D$ of about 0.15 for the data plot the signal in the plot is so weak, that we cannot distinguish the data plot from null plots in a lineup of size $m=20$. \red{The following sentence should be moved back} This pattern is also seen in our experiments as shown in figure \ref{fig:pval_pcorrect}.

>>>>

\begin{table}[htp]
\caption{Possible $p$-values for different numbers of observers, $K$, for fixed size $m = 20$ lineups.}
\begin{center}
\begin{tabular}{|rrr|rrr|rrr|rrr|rrr|}\hline
$K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$ &  $x$ & $p$-value & $K$  & $x$ & $p$-value & $K$  & $x$ & $p$-value\\\hline
1 &  1 & 0.0500 & 2 &  1 & 0.0975 & 3 & 1 & 0.1426 & 4 & 1 & 0.1855 & 5 & 1 & 0.2262 \\%\cline{1-3}
&&& 2 &  2 & 0.0025 & 3 & 2 & 0.0073 & 4 & 2 & 0.0140 & 5 & 2 & 0.0226 \\
&&& &&& 3 & 3 & 0.0001 & 4 & 3 & 0.0005 & 5 & 3 & 0.0012 \\%\cline{1-6}
&&&  &    &        &   &   &        & 4 & 4 & $< 0.0001$ & 5 & 4 & $< 0.0001$ \\\hline
\end{tabular}
\end{center}
\label{pvalue}
\end{table}

@@

\noindent
\red{Note that $F_{X, \theta}$ is the distribution of $X$ under the alternative with given parameter value $\theta$.}

@@

\subsection*{Lineup size $m$}
\noindent
One of the big differences between visual inference and the classical testing situation is the use of a finite number $m-1$ of representatives of the null distribution as comparison against the test statistic. The choice of $m$ has an obvious impact on the test, and we will discuss this in the following paragraph.

The following properties are only derived for the situation of a  fully parameterized  simulation study. This allows us to directly compare lineup tests against their classical counterparts and also identify  properties relevant for a quality assessment of  lineups, which are typically used  in non-parametric situations.

We need to make two crucial assumptions:
\begin{enumerate}
\item  the plot setup is structured in a way that makes it possible for observer to identify a deviation from the null hypothesis,
\item an observer is able to identify the plot with the strongest `signal' (or deviation from $H_0$)  from a lineup.
\end{enumerate}
We will be able to show evidence for the second assumption with the help of the study discussed in section~\ref{sec:simulation}, the first assumption well be reflected by the power of a lineup. The better suited a design is for a particular task, the higher its power will be.

We will now make use of properties of the data going into the lineup and assume that the properties are reflected in the lineup display:

suppose $F_{|t|}(.)$ is the distribution function of an absolute value of $t$, the classical test statistic. Suppose the associated test statistic in a classical test is observed as  $t_{obs}$ with $p$-value $p_D$. By definition we have

@@

%Now suppose $F_{|t|}(.;\delta)$ denotes the distribution function of an absolute value of $t$, the classical test statistic.
%, with non-centrality parameter $\delta$. 
Then the distribution function of the $p$-value $p_D$ under $H_0$ is uniform, since:  
\begin{eqnarray}\label{dist_p}
F{p_D}(p) &=& Pr(p_D \le p)=1-Pr(1-p_D \le 1-p) \nonumber \\
  &=& 1-Pr\left(F_{|t|}^{-1}(1-p_D) \le F_{|t|}^{-1}(1-p) \right) \nonumber \\
  &=& 1-Pr\left(|t_{obs}| \le F_{|t|}^{-1}(1-p)\right) \nonumber \\
  &=&%\left\{ \begin{array}{ll}
          1-F_{|t|}\left( F_{|t|}^{-1}(1-p)\right)=p \mbox{ ; under $H_0$} 
 %         1-F_{|t|}( F_{|t|}^{-1}(1-p); \delta) &\mbox{ ; under $H_1$} 
%       \end{array} \right.     
\end{eqnarray}

@@

Let $p_{0,i}$, $i=1, ..., m-1$ denote the  $p$-values associated with data corresponding to the $m-1$ null plots. Since this data is generated consistently with the null hypothesis,  the $p$-values are independent and  follow a standard Uniform distribution, $p_{i,0} \sim U[0,1], i= 1, ..., m-1$.

@@

With that, we define $Z$  as the number of null plots in a lineup, for which the $p$-value $p_{0,i}$ is smaller than $p_D$.

Then $Z \sim B_{m-1, E[p_D]}$, and the probability that an onlooker will pick the data plot in a given lineup is 
\[
P(Z=0) = \left(1 - E[p_D] \right)^{m-1} = P(p_D \le p_0), \ \ \ \text{ where } p_0 = \min_{1 \le i \le m-1}  \ p_{0,i}.
\]

@@

Figure \ref{fig:pval_power} gives an overview of the probability of picking the data plot for lineups of different size: as $m$ increases we have an increased probability to observe a more highly structured null plot by chance. We see that for a $p$-value $p_D$ of about 0.15 for the data plot the signal in the plot is so weak, that we cannot distinguish the data plot from null plots in a lineup of size $m=20$.  This pattern is confirmed by our findings described in section~ref{sec:simulation}
%also seen in our experiments as shown in figure \ref{fig:pval_pcorrect}.

\red{\#\#\# re-done up to here, HH}
